---
title: "Part VII: MCMC diagnostics for bgms"
author: "Maarten Marsman"
date: "2025-10-03"
format:
  html:
    css:
     - font-awesome.min.css
     - ajohnson.css
    include-in-header: 
      - text: |
          <link rel="stylesheet" href="https://cdnjs.cloudflare.com/ajax/libs/font-awesome/4.7.0/css/font-awesome.min.css">
    code-fold: true
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: false
    df-print: paged
    self-contained: true
engine: knitr
---

::: {.callout-tip title="Workshop Roadmap"}
**You are here:**  

🧭 Intro — From network psychometrics to Bayesian graphical modeling ✅  
📊 **Session I — Quantifying evidence for network edges** ✅  
🔍 **Session II — Modeling topological structure with priors** ✅  
🧠 **Session III — Comparing groups and testing network invariance** ✅   
⚙️ **Session IV — MCMC diagnostics and efficiency** 🟩  
:::

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recap

So far in the workshop, you have learned how to:

- **Estimate and interpret** Bayesian networks within single groups,  
  assessing edge inclusion, parameter uncertainty, and posterior evidence.

- **Model structural assumptions** with different topological priors —  
  Bernoulli, Beta–Bernoulli, and Stochastic Block Models —  
  and understand how these priors shape network density and clustering.

- **Compare groups** by testing invariance and differences,  
  using Bayes factors (Savage–Dickey and model-averaged) to quantify evidence for edge and network-level effects.

---

## Goals for this session

We now turn to **checking the reliability of MCMC output and model adequacy**. At the end of this session, you should be able to:

::: goals
-   **Explain** the difference between bias reduction (warmup) and variance reduction (sampling), and how these are handled in `bgms`.

-   **Compute** and **interpret** MCMC diagnostics:\
    – $\hat{R}$ for convergence,\
    – Effective sample size (ESS) for precision.

-   **Compare** Metropolis vs. NUTS in terms of warmup scheme, efficiency, and diagnostics.

-   **Investigate** diagnostics for selection variables (indicators and pairwise effects).
:::

This session builds directly on your earlier skills:\
- From estimating and comparing networks → to evaluating the quality of the chains that generate those estimates.

::: note
**Recommended Reading:**

-   Margossian, C., & Gelman, A. (2024). For how many iterations should we run Markov chain Monte Carlo? Preprint arXiv:2311.02726v2. doi: 10.48550/arXiv.2311.02726

**Optional Reading**

-   Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P.-C. (2021). Rank-normalization, folding, and localization: An improved $\hat{R}$ for assessing convergence of MCMC. *Bayesian Analysis*, *16*, 667–718. doi: 10.1214/20-BA1221

**R Setup Tip**\
Make sure to install and load the following libraries before running the examples:

``` r
library(bgms)
library(ggplot2)
```
:::

## 1. Motivation

Bayesian network analysis relies on **Markov chain Monte Carlo (MCMC)** to approximate posterior distributions.

A **Markov chain** is a stochastic process where each new state depends only on the current state.\
In **MCMC**, we use such a chain to generate dependent draws. At the start, these draws depend heavily on the initial values.

Over time, if the algorithm is well-behaved, the influence of the starting point *washes out*. The distribution of the draws settles into a stable *pattern* that no longer changes from one step to the next.

This stable long-run distribution is called the **stationary distribution**.

-   For MCMC, in our case, the stationary distribution is the **posterior distribution** we want to approximate.\
-   *Convergence* means that the chain has reached this stationary state.\
-   Only after convergence can we use the draws to summarize the posterior.

---

Unlike direct simulation (e.g., using `rnorm` or `rbinom`), MCMC does not give independent draws from the posterior.

MCMC samples are:\
- **Dependent** — each draw is generated from the previous state of the chain.\
- **Approximate** — the chain only approaches the true posterior after some iterations.

This raises two key questions:\
1. **Convergence (bias):** Has the chain reached the stationary posterior distribution?\
2. **Precision (variance):** Have we collected enough *effectively independent* samples to summarize the posterior reliably?

In short: *Did the chain converge?* and *Do we have enough information after convergence?*

------------------------------------------------------------------------

## Assignment: Match Traceplots to Posteriors

The figure below shows three traceplots (A–C) and three posterior distributions (1–3).\
Each traceplot comes from a different Markov chain, and each posterior is the distribution estimated from that chain.

**Your task:** Match each traceplot to its corresponding posterior distribution.

```{r}
set.seed(42)

# True posterior: Beta(7,17)
alpha <- 7; beta <- 17
n_draws <- 2000

# --- Chains ---
# Scenario 1: Autocorrelated + bad start
chain1 <- numeric(n_draws)
chain1[1] <- 0.9
for (i in 2:n_draws) {
  chain1[i] <- chain1[i-1] + rnorm(1, 0, 0.01)
  chain1[i] <- min(max(chain1[i], 0), 1)
}

# Scenario 2: Perfect
chain2 <- rbeta(n_draws, alpha, beta)

# Scenario 3: Drift
chain3 <- rbeta(n_draws, alpha, beta) + seq(0, 0.5, length.out=n_draws)
chain3 <- pmin(chain3, 1)

# Densities
dens1 <- density(chain1, from=0, to=1)
dens2 <- density(chain2, from=0, to=1)
dens3 <- density(chain3, from=0, to=1)

# True posterior
true_x <- seq(0, 1, length=200)
true_y <- dbeta(true_x, alpha, beta)

# Layout
layout(matrix(c(1,4,2,5,3,6), ncol=2, byrow=TRUE))
par(cex=1, mar=c(4,4,2,1))

# --- Traceplots ---
plot(chain1, type="l", col="steelblue", main="Traceplot A",
     xlab="", ylab="", axes=FALSE, ylim = c(0,1))
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("Number of Iterations", side=1, line=2.5, cex = 1.7)
mtext("Sampled Value", side=2, line=2.5, cex = 1.7)

plot(chain2, type="l", col="steelblue", main="Traceplot B",
     xlab="", ylab="", axes=FALSE, ylim = c(0,1))
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("Number of Iterations", side=1, line=2.5, cex = 1.7)
mtext("Sampled Value", side=2, line=2.5, cex = 1.7)

plot(chain3, type="l", col="steelblue", main="Traceplot C",
     xlab="", ylab="", axes=FALSE, ylim = c(0,1))
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("Number of Iterations", side=1, line=2.5, cex = 1.7)
mtext("Sampled Value", side=2, line=2.5, cex = 1.7)

# --- Posterior densities ---
plot(true_x, true_y, type="l", lwd=3, col="black",
     main="Posterior 1", xlab="", ylab="", axes=FALSE, ylim=c(0,12))
lines(dens3, col="orange", lwd=2)
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("theta", side=1, line=2.5, cex = 1.7)
mtext("Density", side=2, line=2.5, cex = 1.7)

plot(true_x, true_y, type="l", lwd=3, col="black",
     main="Posterior 2", xlab="", ylab="", axes=FALSE, ylim=c(0,12))
lines(dens1, col="red", lwd=2)
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("theta", side=1, line=2.5, cex = 1.7)
mtext("Density", side=2, line=2.5, cex = 1.7)

plot(true_x, true_y, type="l", lwd=3, col="black",
     main="Posterior 3", xlab="", ylab="", axes=FALSE, ylim=c(0,12))
lines(dens2, col="forestgreen", lwd=2)
axis(1, cex.axis=1.2); axis(2, las=1, cex.axis=1.2)
mtext("theta", side=1, line=2.5, cex = 1.7)
mtext("Density", side=2, line=2.5, cex = 1.7)
```

::: panel-tabset
### Tasks

Look at the three traceplots and their corresponding posterior densities.

1. Which chain still looks influenced by where it started?  
2. Which chain shows a systematic drift over time?  
3. Which chain appears to have settled well into the target distribution?  
4. How do these behaviors show up in the estimated posteriors compared to the true posterior (black curve)?

### Try it

Discuss in small groups for about 5–10 minutes.  
Focus on:  
- Does each chain explore the full plausible range or remain stuck or drifting?  
- Do the posterior estimates align with the black curve (the true posterior)?  
- What would these patterns mean for interpreting results?

### Solution

**Traceplot B → Posterior 3:** Well-behaved chain; explores freely, posterior matches the truth.  
**Traceplot A → Posterior 2:** Still influenced by its starting point; moves slowly, posterior biased.  
**Traceplot C → Posterior 1:** Systematic drift; posterior shifted away from the truth.

**Take-away messages**  
- A chain must both **forget its starting point** (warmup/convergence) and **move freely** around the posterior (good mixing).  
- If either condition fails, posterior inference becomes unreliable.
:::

------------------------------------------------------------------------

### Warmup vs. Sampling

MCMC sampling has two distinct phases:

-   **Warmup (burn-in):**\
    Early iterations are used to adapt the sampler and move the chain toward the stationary posterior.\
    These draws are discarded because they still reflect the starting values.

-   **Sampling phase:**\
    Once adaptation has stabilized, subsequent iterations are kept for inference.

    -   **Number of samples:** more samples generally mean more precise estimates.\
    -   **Effective sample size (ESS):** accounts for the dependence between MCMC draws.\
        ESS tells us how many independent samples would provide the same amount of information as the dependent MCMC draws.

------------------------------------------------------------------------

### Diagnostics

-   $\mathbf{\hat{R}}$ (potential scale reduction statistic):\
    Compares variation *within chains* to variation *between chains*.
    -   $\hat{R} \approx 1$: chains have converged to the same distribution.\
    -   In practice, $\hat{R} < 1.01$ (Vehtari et al., 2021) is recommended as the threshold.
-   **Effective sample size (ESS):**\
    A statistic that measures the *information content* of the chain.
    -   Larger ESS → more reliable, precise estimates.\
    -   Smaller ESS → more noisy estimates.

------------------------------------------------------------------------

## Assignment: Inspecting Chains in `bgms`

Let’s see how these ideas look in practice with `bgms`. We will fit a model **without edge selection** (`edge_selection = FALSE`), so the sampler only estimates parameters.

::: panel-tabset
### Tasks

1. **Fit the model**  
   Run `bgm()` on the `ADHD` dataset with `edge_selection = FALSE`.  
   This estimates the network without selection so we can inspect MCMC behavior.

2. **Inspect traceplots**  
   Produce traceplots for a few parameters.  
   Check by eye whether the chains move freely and mix well across the full range of plausible values.

3. **Check diagnostics**  
   Use the summary output to view $\hat{R}$ and effective sample size (ESS) values.  
   Recall:  
   - $\hat{R}$ close to 1 indicates good convergence across chains.  
   - Larger ESS means more precise posterior estimates.

4. **Evaluate**  
   - Do the chains look stationary after warmup?  
   - Do $\hat{R}$ values suggest convergence?  
   - Are ESS values large enough for stable inference?

### Try it

Fit a `bgm` model on the ADHD dataset with `edge_selection = FALSE`. Then:

1.  Produce **traceplots** for 2–4 parameters.\
2.  Run `summary(fit)` and inspect the reported $\hat{R}$ and ESS values.

Use the outputs to answer these questions: - Do the traceplots suggest the chains have converged?\
- Are the $\hat{R}$ values close to 1?\
- Are the effective sample sizes large enough to trust the estimates?

Here is some example code to scaffold your work:

```{r, eval = FALSE}
library(bgms)

# Fit model without edge selection
fit = bgm(ADHD, edge_selection = FALSE)

# Inspect convergence diagnostics
summary(fit)

# Traceplots for some parameters
main_samples = do.call(rbind, fit$raw_samples$main)
colnames(main_samples) = fit$raw_samples$parameter_names$main
pairwise_samples = do.call(rbind, fit$raw_samples$pairwise)
colnames(pairwise_samples) = fit$raw_samples$parameter_names$pairwise

plot(pairwise_samples[, 1], type = "l")
```

### Solution

-   **Traceplots**\
    Well-behaved chains should look like “fat hairy caterpillars” that wander freely across the posterior range.\
    If chains stay stuck, drift over time, or show very different ranges, that suggests non-convergence.

-   $\hat{R}$ values\
    Numbers very close to 1 (typically \< 1.01) indicate that the chains are sampling from the same distribution.\
    If $\hat{R}$ is noticeably larger than 1, convergence is doubtful.

-   **Effective sample sizes (ESS)**\
    Larger ESS means more information and smaller Monte Carlo error.\
    Small ESS indicates the estimates are less precise and may not be trustworthy.
:::

------------------------------------------------------------------------

**Takeaway:**\
- Warmup reduces **bias**.\
- The number of samples controls **precision**.\
- ESS tells us how informative those samples actually are.\
- $\hat{R}$ tells us whether chains have converged.

------------------------------------------------------------------------

### What if diagnostics are not satisfactory?

-   **If** $\hat{R}$ is larger than 1.01\
    → Chains have not mixed to the same distribution.
    -   Increase the number of iterations.\
    -   Run more chains with dispersed starting values.\
    -   Consider switching to a more efficient sampler (e.g. NUTS instead of Metropolis).
-   **If ESS is small**\
    → The effective number of independent samples is too low for precise inference.
    -   Increase the number of iterations.\
    -   Reduce autocorrelation by using a more efficient sampler.

------------------------------------------------------------------------

## 2. MCMC Workflow & Control Parameters

Every MCMC run follows a **workflow**:\
1. **Initialization** – set starting values for each chain.\
2. **Warmup (burn-in)** – adapt tuning parameters and move the chain toward the stationary posterior (draws are discarded).\
3. **Sampling** – collect draws for inference once adaptation has finished.\
4. **Diagnostics** – evaluate convergence ($\hat{R}$) and precision (ESS).

How this workflow is carried out depends on the **control parameters** we choose:\
- number of chains,\
- number of iterations,\
- length of warmup,\
- and the type of update method (e.g., Metropolis or NUTS).

------------------------------------------------------------------------

### Stan-style Warmup Schedule in `bgms` for HMC and NUTS methods

Warmup for HMC/NUTS in `bgms` is structured into **stages**:

1.  **Stage 1 – Fast adaptation (\~7.5% of warmup)**
    -   Adapt the step size quickly so the sampler begins to explore.
2.  **Stage 2 – Adaptive windows**
    -   Build doubling “windows” of increasing length (25, 50, ...).\
    -   Within each window:
        -   (Optional) Update the **mass matrix** to match posterior curvature.\
        -   Refine the step size using dual averaging.
3.  **Stage 3a – Final stabilization (\~last 10%)**
    -   Stabilize the step size.\
    -   Lock in the final mass matrix if estimated.
4.  **Stage 3b – Edge selection adaptation (if enabled)**
    -   Adapt the **proposal standard deviations** for Metropolis edge moves.
5.  **Stage 3c – Final adjustments (if enabled)**
    -   Make final adjustments of the global step size after edge adaptation.

------------------------------------------------------------------------

After these stages, **warmup ends** and the sampler enters the **sampling phase**.

### Warmup for Metropolis (Random Walk proposals) in `bgms`

In `bgms`, warmup for Metropolis-within-Gibbs is much simpler than for NUTS:

-   **Adaptation goal:** learn suitable proposal standard deviations (step sizes) for the random walk updates.\
-   **How:** use a Robbins–Monro adaptation scheme.
    -   After each iteration, acceptance probabilities are used to adjust proposal scales.\
    -   The update is weighted by a decaying factor (∝ iteration\^(-0.75)) so adaptation gradually stabilizes.\
    -   Target acceptance rate is around 0.44 (standard for one-dimensional random walks).
-   **End of warmup:** once the proposal scales have adapted, the sampler fixes them and begins the sampling phase.

------------------------------------------------------------------------

**Key difference from warmup for NUTS:**\
- NUTS warmup involves multiple stages (step size tuning, mass matrix adaptation, final stabilization).\
- Metropolis warmup is a single-stage process focused only on proposal standard deviations.

In both cases, **warmup draws are discarded** because they reflect adaptation rather than stationary sampling.

------------------------------------------------------------------------

## Assignment: Relative efficiency MCMC procedures

We now examine how these choices **show up in diagnostics**. The next exercises compare samplers and control parameters in terms of $\hat{R}$ (convergence) and ESS (information).

::: panel-tabset
### Tasks

1. **Fit the ADHD model twice**  
   Estimate the model with `edge_selection = FALSE` using two update methods:  
   - `update_method = "metropolis"`  
   - `update_method = "nuts"`  
   This allows you to compare how the samplers differ in convergence and efficiency.

2. **Inspect diagnostics**  
   Compare $\hat{R}$ and ESS values in the model summaries.  
   Check whether both samplers achieve good convergence ( $\hat{R} \approx 1$ ) and note any differences in effective sample sizes.

3. **Examine traceplots**  
   Create traceplots for a few parameters from each sampler.  
   Assess visually how quickly and smoothly each chain explores the posterior.

4. **Evaluate**  
   - Do both samplers appear to have converged?  
   - Which produces higher ESS for the same number of iterations?  
   - How do the traceplots differ in terms of mixing speed and stability?


### Try it

Run both samplers with a modest number of iterations (e.g., `iter = 1000`) so that differences are easy to spot.

1.  Fit the model once with **Metropolis** and once with **NUTS** (`edge_selection = FALSE`).\
2.  Call `summary(fit)` for each run and compare the reported $\hat{R}$ and ESS values.\
3.  Produce traceplots for 2–4 parameters.

Use these outputs to answer:\
- Do the chains appear to have converged?\
- Are the $\hat{R}$ values close to 1 for both samplers?\
- Which sampler provides larger ESS for the same number of iterations?

```{r, eval = FALSE}
library(bgms)

fit_met <- bgm(ADHD,
               update_method   = "adaptive-metropolis",
               edge_selection  = FALSE,
               iter = 1000, 
               seed = 4321)

fit_nuts <- bgm(ADHD,
                update_method   = "nuts",
                edge_selection  = FALSE,
                seed = 4321)

summary(fit_met)
summary(fit_nuts)
```

### Solution

-   **Traceplots**
    -   Metropolis chains often move slowly at 1,000 iterations and may not explore the full posterior range.\
    -   NUTS chains typically mix more quickly and cover the posterior more efficiently.
-   $\hat{R}$
    -   For NUTS, values are usually close to 1.\
    -   For Metropolis, some parameters may have $\hat{R} > 1.01$, indicating incomplete convergence.
-   **ESS**
    -   NUTS generally achieves a larger ESS at the same number of iterations.\
    -   Metropolis requires more iterations to reach comparable ESS.

**Takeaway message**\
With short runs, NUTS is usually more efficient, showing better convergence ($\hat{R}$) and higher information content (ESS).\
Metropolis can still reach satisfactory diagnostics, but only with longer chains or more careful control parameters.
:::

## Assignment B: Improving Metropolis via control parameters

In the previous exercise, we saw that Metropolis underperformed at 1,000 iterations.\
The goal here is to show that Metropolis diagnostics improve when we **increase the number of iterations and/or the number of chains**.

::: panel-tabset
### Tasks

Building on the previous exercise (where Metropolis underperformed at 1,000 iterations), now explore how diagnostics improve when you adjust the control parameters.

1. **Refit the model**  
   Run the Metropolis sampler again (`edge_selection = FALSE`), focusing on how longer runs affect convergence.

2. **Increase iterations**  
   Try increasing the total number of iterations (e.g., from 1,000 → 4,000 or 8,000).  
   Optionally, increase the number of chains (e.g., from 4 → 6).

3. **Inspect diagnostics**  
   Compare $\hat{R}$ and ESS across runs.  
   Note whether these values improve as you increase sampling length or the number of chains.

4. **Check traceplots**  
   Examine whether chains now look stable, overlapping, and well mixed.  
   Do they explore the full range of the posterior?

5. **Evaluate**  
   - Does $\hat{R}$ approach 1 as iterations or chains increase?  
   - Does ESS rise accordingly?  
   - At what settings do diagnostics become satisfactory?

### Try it

In the previous exercise, Metropolis did not perform well at 1,000 iterations.\
Now rerun the model with **Metropolis only** and adjust the control parameters:

1.  Increase the number of iterations (e.g., from 1,000 → 2,000 or 10,000).\
2.  Optionally increase the number of chains (e.g., from 4 → 8).\
3.  Inspect the $\hat{R}$ and ESS values in the summaries.

Reflect on:\
- Does $\hat{R}$ move closer to 1 as iterations and/or chains increase?\
- Does ESS grow with longer runs?\
- At what settings would you consider the diagnostics satisfactory?

```{r, eval = FALSE}
# Metropolis: longer chains
fit_met_more_iter <- bgm(ADHD,
                         update_method = "adaptive-metropolis",
                         edge_selection = FALSE,
                         chains = 4, 
                         iter = 2000, 
                         seed = 4321)
                  

# Metropolis: more chains
fit_met_more_chains <- bgm(ADHD,
                           update_method = "adaptive-metropolis",
                           edge_selection = FALSE,
                           chains = 8, 
                           iter = 1000, 
                           seed = 4321)
                         

summary(fit_met_more_iter)
summary(fit_met_more_chains)
```

### Solution

-   $\hat{R}$\
    Moves closer to 1 as the number of iterations increases and when more chains are added.

-   **ESS**\
    Grows with longer runs; more iterations generally produce larger ESS values.

-   **Traceplots**\
    With more iterations, chains tend to overlap and appear stationary, indicating improved mixing.

**Takeaway message**\
Metropolis can achieve satisfactory diagnostics when control parameters are scaled appropriately.\
Increasing iterations (and, when useful, chains) improves $\hat{R}$ and ESS, allowing Metropolis to approach the reliability of NUTS given sufficient sampling.
:::

------------------------------------------------------------------------

### From general diagnostics to variable selection

So far, we have studied MCMC diagnostics in general:\
- Traceplots as a visual tool,\
- $\hat{R}$ for convergence,\
- ESS as a measure of information in dependent draws.

But in `bgms`, our main interest is usually in **edge selection**:\
- **Binary indicators** for edge inclusion/exclusion,\
- **Spike–and–slab parameters**, e.g., the pairwise interactions, or their group differences, that are being selected by the indicators.

```{r}
library(ggplot2)

# Parameters
p <- 0.7   # inclusion probability (weight on slab)

# Grid
x <- seq(-4, 4, length.out = 500)

# Slab density scaled by p
slab <- p * dnorm(x, mean = 0, sd = 1)

# Data for slab
df_slab <- data.frame(x = x, density = slab)

# Data for spike
df_spike <- data.frame(x = 0, xend = 0, y = 0, yend = 1 - p)

# Plot
ggplot(df_slab, aes(x = x, y = density)) +
  geom_line(color = "#D55E00", linewidth = 1.2) +
  geom_segment(data = df_spike,
               aes(x = x, xend = xend, y = y, yend = yend),
               inherit.aes = FALSE,
               color = "#0072B2", linewidth = 1.4, lineend = "butt") +
  labs(x = "Parameter value", y = "Density", 
       title = "Spike-and-Slab Prior") +
  annotate("text", x = 0.1, y = 1 - p + 0.02, label = "Spike", color = "#0072B2", hjust = 0, size =6) +
  annotate("text", x = 2, y = max(slab), label = "Slab", color = "#D55E00", hjust = 0,size=6) +
  theme_minimal(base_size = 14)
```



To evaluate MCMC samples of them properly, we need to understand how $\hat{R}$ and ESS apply in this setting, and how we can derive effective sample sizes for these discrete or mixture-valued parameters.

------------------------------------------------------------------------

## 3. Diagnostics for Variable Selection Parameters

So far, we have treated ESS and $\hat{R}$ in general.  
In `bgms`, however, we care about **variable selection parameters**:

- **Indicators** behave like a two-state Markov chain.  
  - The chain can get “stuck” in 0 or 1 for long stretches.  
  - This produces **high autocorrelation** $\Rightarrow$ low ESS.  

- **Spike–and–slab parameters** are a mixture of a point mass (spike at zero when not selected) and a continuous slab (a density when selected).  
  - Switching between spike and slab creates additional autocorrelation.  
  - ESS can therefore often be low for these parameters as well.  

- **Takeaway:** Even when $\hat{R} \approx 1$, ESS for variable selection parameters can be modest. This affects the *precision* of reported edge inclusion probabilities.  

---

### ESS for binary indicators

<div class="describe">

**Closed-form ESS for binary indicators**

For a binary indicator $\gamma_t$, modelled as a two-state Markov chain with 
transition probabilities $\hat{a} = \Pr(0 \to 1)$ and $\hat{b} = \Pr(1 \to 0)$:

- Lag-1 autocorrelation: $\rho_1 = 1 - (\hat{a}+\hat{b})$.  
- Autocorrelations decay geometrically: $\rho_k = \rho_1^k$.  
- Integrated autocorrelation time:
$$
\tau_{\text{int}} = \frac{2 - (\hat{a}+\hat{b})}{\hat{a}+\hat{b}} .
$$
- Effective sample size:
$$
\mathrm{ESS}(\gamma) = \frac{T}{\tau_{\text{int}}}.
$$
</div>

---

### ESS for spike–and–slab parameters

<div class="describe">

**Delta method derivation for $\hat{\theta} = \hat{p}\cdot\hat{\mu}$**

- Let $\hat{p}$ = posterior inclusion probability (from the indicator chain).  
- Let $\hat{\mu}$ = conditional mean given inclusion.  
- Our estimator of the marginal mean is
$$
\hat{\theta} = \hat{p}\,\hat{\mu}.
$$

By the delta method, treating $\hat{p}$ and $\hat{\mu}$ as approximately uncorrelated:

$$
\operatorname{Var}(\hat{\theta})
\approx \hat{\mu}^2 \,\operatorname{Var}(\hat{p})
+ \hat{p}^2 \,\operatorname{Var}(\hat{\mu}).
$$

Thus the Monte Carlo standard error is

$$
\text{MCSE}(\hat{\theta}) =
\sqrt{\hat{\mu}^2\,\text{MCSE}^2(\hat{p}) +
      \hat{p}^2\,\text{MCSE}^2(\hat{\mu})}.
$$

Finally, since
$$
\text{MCSE}^2 = \frac{\text{Posterior Variance}}{\text{ESS}},
$$
we can solve for an **effective sample size for $\hat{\theta}$**:

$$
\text{ESS}_{\text{combined}} =
\frac{\operatorname{Var}(\theta)}{\operatorname{Var}(\hat{\theta})}.
$$
</div>


## Assignment: Investigate $\hat{R}$ and ESS for Selection Variables

We now analyze the ADHD dataset **with edge selection** using `update_method = "nuts"`. The goal is to compare diagnostics for **binary indicators** and **spike–and–slab parameters** (i.e., the pairwise interactions).

::: {.panel-tabset}

### Tasks

1. **Fit the model**  
   Estimate a `bgm` model on the `ADHD` dataset using  
   `update_method = "nuts"` and `edge_selection = TRUE`, running multiple chains.

2. **Inspect the summary**  
   For several edges, compare diagnostics ($\hat{R}$, ESS) for both the **binary indicators** (edge inclusion) and the corresponding **spike–and–slab parameters** (pairwise interactions).  
   Note their posterior means.

3. **Visualize chain behavior**  
   Create traceplots for the same edges to check whether the chains mix well and explore the posterior adequately.

4. **Compare across parameter types**  
   - Are $\hat{R}$ values acceptable for both indicators and slabs?  
   - Are ESS values systematically lower for selection variables than for other continuous parameters?  
   - How do these diagnostics compare with the ESS values you obtained earlier *without* selection?

Reflect on what these patterns imply about the additional sampling difficulty introduced by discrete selection variables.


### Try it

1. Fit `bgm` with `update_method = "nuts"` and `edge_selection = TRUE` (e.g., `chains = 4`, a few thousand iterations).  
2. Pull the two diagnostics tables:
   - `summary(fit)$indicator`  → **Indicators** (edge inclusion).  
   - `summary(fit)$pairwise`   → **Pairwise effects** (interaction parameters).  
3. Pick 3–5 **matched edges** (same edge in both tables). Include
   - at least one edge with inclusion ≈ 0 or ≈ 1 (to see constant indicators/NA diagnostics), and  
   - at least one edge with inclusion in the middle (e.g., 0.2–0.8) to see switching.  
4. Compare the diagnostics columns for those edges:
   - `Rhat`, `n_eff` (ESS), and `mcse` (if reported).  
5. Make **traceplots** for the same edges (both the indicator and the pairwise effect) and relate the plots to the numbers.

**Guiding questions**
- Which edges have the **lowest ESS** for **indicators**? For **pairwise effects**?  
- Do the traceplots explain the ESS values (e.g., slow switching for pairwise effects)?  
- Are the **$\hat{R}$** values acceptable (≈ 1), even when ESS is modest?  
- Do any indicators show **constant 0 or 1** (hence NA diagnostics)? How does that relate to their pairwise effects?  
- How do these diagnostics compare to your earlier run **without selection**?

```{r, eval = FALSE}
library(bgms)
#no selection
fit_nosel = bgm(ADHD,
                update_method = "nuts",
                edge_selection = FALSE,
                chains = 4, 
                iter = 1000, 
                seed = 2025)
#selection
fit_sel = bgm(ADHD,
              update_method = "nuts",
              edge_selection = TRUE,
              chains = 4, 
              iter = 1000, 
              seed = 2025)

#summaries pairwise effects
summary(fit_nosel)$pairwise
summary(fit_sel)$pairwise

#summaries indicators
summary(fit_sel)$indicator
```
### Solution

- **$\hat{R}$:** With NUTS and sufficient iterations, $\hat{R} \approx 1$ for both indicators and pairwise effects, showing convergence across chains.  

- **ESS:**  
  - Indicators can show **lower ESS** because they are two–state variables and the chain can dwell in 0 or 1 for long stretches.  
  - Pairwise effects also can have **modest ESS**, reflecting the slow switching between spike and slab.  
  - Continuous parameters such as thresholds usually achieve much higher ESS under the same run.  

- **Traceplots:**  
  - Pairwise effects also show sequences of exact zeros whenever the indicator is 0, 
    and continuous variation (slab values) when the indicator is 1.
    
**Takeaway**  
Even with an efficient sampler (NUTS) and acceptable $\hat{R}$, **ESS for indicators and pairwise effects can be modest**.  
When interpreting results, always check MCSE alongside $\hat{R}$ and ESS, and if precision is insufficient, increase iterations or chains.  

:::
### Closing

A key feature of `bgms` is that it reports $\hat{R}$ and ESS not only for continuous parameters,  
but also for **selection variables** (indicators and pairwise effects).  

As far as we know, this is the only implementation that provides full diagnostics  
for these discrete and mixture-valued parameters. This makes it possible to:

- Evaluate convergence and mixing of selection variables,  
- Quantify how much effective information underlies edge inclusion probabilities,  
- Compare diagnostics across indicators, pairwise effects, and continuous parameters.  

---

### Take-home message

This completes the **Bayesian workflow** for network models:

1. **Estimate** the model.  
2. **Diagnose** the chains — check $\hat{R}$, ESS, and MCSE for all parameters,  
   including indicators and pairwise effects.  
3. **Interpret** results only after diagnostics confirm reliability.

With this final step, you have completed the full Bayesian workflow for graphical models —  
from prior specification and model estimation to inference, comparison, and validation.  

---

### Back to the workshop goals

At the start of the workshop, we set out to:

- **Explain** the benefits of a Bayesian approach to network analysis.  
- **Estimate and interpret** network models using R and JASP.  
- **Understand** how priors affect network structure and complexity.  
- **Examine** group differences using Bayes factors.  
- **Evaluate** convergence and precision with MCMC diagnostics.  

---

### Closing

Thank you for participating in the workshop.  

We hope these sessions have given you both conceptual insight and practical skills for applying Bayesian methods to graphical models.  
If you have questions or would like to discuss applications of these techniques, please feel free to reach out!


